---
title: "Limited Dependent Variables"
subtitle: "Demand Estimation"
author: "Jose M. Fernandez"
institute: "University of Louisville"
date: "2020-7-4 (updated: `r Sys.Date()`)"
output:
  xaringan::moon_reader:
    css: [default, uol, uol-fonts, "extra.css"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
class: inverse, center, middle
# Demand Estimation with aggregate market shares
---
## Micro versus Market level data
One of the advantages of having microdata is that you can estimate demand directly from individual consumer choices.

But often times, we do not get to observe individual transaction data. 

Instead, we observe market level data where we know the market share for a certain product and product characteristics. 

---
## Another problem: price endogeneity?

Note: in deriving all these examples, an implicit assumption is that the distribution of the $\epsilon_{ij}$'s are independent of the prices. This is analogous to assuming that prices are exogenous.

Case study: Trajtenberg (1989) study of demand for CAT scanners. Disturbing finding: coefficient on price is positive, implying that people prefer more expensive machines!

Possible explanation: quality differentials across products not adequately controlled for. In differentiated product markets, where each product is valued on the basis of its characteristics, brands with highly-desired characteristics (higher quality) may command higher prices. If any of these characteristics are not observed, and hence not controlled for, we can have endogeneity problems. $E[p\epsilon] \ne 0$

---
## Estimation with aggregate market shares

Next we consider how to estimate demand functions in the presence of price endogeneity, and when the researcher only has access to aggregate market shares. 

This summarizes findings from Berry (1994).

---
## Data

Our data for a particular market looks like this

j | $\widehat{s}_{mj}$ | $p_j$ | $X_{1j}$ | $X_{2j}$
---|---------|-------|------|------
A | 25% | $1.50 | red | large
B | 35% | $2.00 | blue | small
C | 45% | $2.50 | green | large

Total Market size = $M$
Total number of brands = $J$

We want to use these data to estimate demand for different products using differences in market share and characteristics across different brands and different markets.
---
## Model

Let a consumer's utility function be $$U_{ijm}=X_{jm}\beta-\alpha p_{jm}+\xi_{jm}+\epsilon_{ijm}$$
where i indexes consumer, j indexes product, and m indexes market.

Let $\delta_{jm}=X_{jm}\beta-\alpha p_{jm}+\xi_{jm}$

Econometrician observes neither $\xi_{jm}$ or $\epsilon_{ij}$, but household i observes both.

You can think of $\xi_{jm}$ as a product specific unobserved quality shock. It is easy to see that high quality products (high $\xi$) would imply consumers are willing to pay more for the good so price would be higher as well. $E[\xi p \neq 0]$

The other error is an idiosyncratic shock, which we will assume is distributed Type I Extreme Value across all consumers, brands, and markets.
---
## Probabilities

Given our assumption of the idiosyncratic error, we can write the choice probabilities as a conditional logit probability.

$$Pr(y_{ijm}=j)=\frac{exp(\delta_{jm})}{\sum_{k=1}^J exp(\delta_{km})}$$
---
## Predicted market shares

We can't estimate this probability directly, but we do observe the actual market shares. So we can transform this probability into a _predicted market share_

That is, if we knew the values of the $\delta$'s, then we could construct the actual market shares. 

### Predicted market share

$$\widetilde{s}_{jm}=Pr(y_{ijm}=j)=\frac{exp(\delta_{jm})}{\sum_{k=1}^J exp(\delta_{km})}$$

We need to normalize one product such that $\delta_{0m}=0$

$$\widetilde{s}_{0m}=Pr(y_{ijm}=0)=\frac{1}{1+\sum_k exp(\delta_{km})}$$
---
## Transform shares

We can then use these predicted shares to make a linear equation by taking logs

$$log(\widetilde{s}_{jm})-log(\widetilde{s}_{0m})=\delta_{jm}=X_{jm}\beta-\alpha p_{jm}+\xi_{jm}$$
__Construct our new dependent variable__

Since we actually observe the market shares of each product in each market, we can construct our actual values of $\delta_{jm}$.

Let $s_{jm}$ be the actual market share of product j in market m.

Then the actual mean utility is  $\widehat{\delta}_{jm}=log({s}_{jm})-log({s}_{0m})$
---
## Objective Function

We calculate this for every market. Our objective function becomes 
$$\begin{align*}E[\xi Z]&=\frac{1}{MM}\frac{1}{J}\sum_{m=1}^M \sum_{j=1}^J \xi_{jm}Z_{jm} \\ \\ 
&=\frac{1}{MM}\frac{1}{J}\sum_{m=1}^M \sum_{j=1}^J [\widehat{\delta}_{jm}-X_{jm}\beta-\alpha p_{jm}]Z_{jm}\end{align*}$$

Where Z is a set of instrumental variables and MM is the total number of markets.

Basically, we reduce the whole problem down to an instrumentals variable problem.
---
## What are appropriate instruments

* Cost shifters (prices of raw materials or workers; industry dependent)
* Characteristics of competitors' products
* If panel, the price of the same good, but in other markets

---
## Berry 1994 in R: Data

```{r warning=FALSE, message=FALSE}
library(AER)
blpdata<-hdm::BLP$BLP #BLP Data
blpZ<-hdm::BLP$Z #Instrumental Variables
blpZ<-as.data.frame(blpZ)
blp.var.names=c("model name","model id","firm id","cdid",
                "log price","miles per gallon","miles per dollar",
                "horse power per weight","air conditioning",
                "size of the car","market share","outside option share",
                "log share j - log share 0","time trend")
ols.1<-lm(y~price+hpwt+space+mpg+air+mpd+factor(model.id), data=blpdata)
iv.1<-ivreg(y~price+hpwt+space+mpg+air+mpd+factor(model.id)|hpwt+space+mpg
            +air+mpd+factor(model.id)+blpZ$sum.other.1+blpZ$sum.other.hpwt
            +blpZ$sum.other.air+blpZ$sum.other.mpd
            +blpZ$sum.other.space+blpZ$sum.rival.1+blpZ$sum.rival.hpwt
            +blpZ$sum.rival.air+blpZ$sum.rival.mpd+blpZ$sum.rival.space, data=blpdata)
```
---
## Berry 1994 in R: Results 
<small>
```{r echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
stargazer::stargazer(ols.1,iv.1,type="html",covariate.labels=c("price","Horse power per wgt","size of car","miles per gallon","air conditioning","miles per dollar"),omit = "model.id",dep.var.labels.include = FALSE, font.size = "small", single.row = TRUE, colnames = FALSE, style = "aer", model.numbers = FALSE, model.names = FALSE, column.labels = c("OLS","IV"), omit.stat = c("adj.rsq","ser"), star.char = c("a","b","c"))
```
</small>
