---
title: "Limited Dependent Variables"
subtitle: "Count Models"
author: "Jose M. Fernandez"
institute: "University of Louisville"
date: "2020-7-4 (updated: `r Sys.Date()`)"
output:
  xaringan::moon_reader:
    css: [default, uol, uol-fonts, "extra.css"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

## Count Models

In these notes, we will discuss several types of count models

- Ordered Probit/Logit
- Poisson Model
- Negative Binomial

---
class: inverse, middle, center
# Ordered Probit/Logit
---
## Oridinal Categorical Variables

Suppose your dependent is categorical in nature and these categories show an ordinal proper (i.e. you can order them)

Examples: Surveys including Likert Scales (strongly agree, agree, neither, disagree, strongly disagree)

Medalling in the olympics (Gold, Silver, Bronze)

Soda size preference (small, medium, large, extra large)
---
## Ordered Probit/Logit

In these cases of ordinal categorical variables, we use an ordered probit/logit model.

Notice, these responses, while they can be ordered, the distance between levels is not constant. 

That is, we do not know what increase is necessary to move someone from _strongly disagree_ to just _disagree_.
---
## Ordered Probit/Logit Math

The math of the ordered probit/logit is very similar to a simple binary model

There is just one difference, we need to estimate level parameters.
---
## Ordered Probit/Logit Math

The math to the ordered probit/logit is very similar to a simple binary model

There is just one difference, we need to estimate level parameters.

Consider three observed outcomes: y = 0,1,2.
Consider the latent variable model without a constant:
$$y^{*} = \beta_1 x_1 +...+\beta_k x_k +\epsilon$$
where $\epsilon \rightarrow N(0,1)$. Define two cut-off points: $\alpha_1 < \alpha_2$
We do not observe $y^{*}$, but we do observe choices according to the following rule

- y = 0 if $y^{*} < \alpha_1$;
- y = 1 if $\alpha_1 < y^{*} < \alpha_2$;     
- y = 2 if $\alpha_2 < y^{*}$

---
## Analytical Example

y = 0 if inactive, y = 1 if part-time, and y = 2 if full-time

$y* = \beta_e educ +\beta_k kids +\epsilon$, where $\epsilon \rightarrow N (0,1)$

Then
y = 0 if $\beta_e  educ +\beta_k  kids +\epsilon$ < $\alpha_1$

y = 1 if $\alpha_1 < \beta_e educ +\beta_k kids +\epsilon$ < $\alpha_2$

y = 2 if $\alpha_2 < \beta_e  educ +\beta_k kids +\epsilon$

We could alternatively introduce a constant $\beta_0$ and assume that $\alpha_1$ = 0.

---
## Log Likelihood Function

$$Log \, L=\sum_{i=1}^{n} \sum_j Z_{ij}[F_{ij}-F_{ij-1}]$$

where $F_{ij}=F(\alpha_j-\beta_e  educ -\beta_k  kids)$ and $Z_{ij}$ is an indicator that equals 1 if person i select option j and zero otherwise.

We choose the values of $\alpha$ and $\beta$ that maximize the value of the log likelihood function.

__However, in the case of models such as ordered probit and ordered logit failure to account for heteroskedasticity can lead to biased parameter estimates in addition to misspecified standard errors__

---
## Ordered Logit in R

A study looks at factors that influence the decision of whether to apply to graduate school. 

College juniors are asked if they are unlikely, somewhat likely, or very likely to apply to graduate school. 

Hence, our outcome variable has three categories. 

Data on parental educational status, whether the undergraduate institution is public or private, and current GPA is also collected.

---
## Example: Data
```{r}
library(foreign)
dat <- read.dta("https://stats.idre.ucla.edu/stat/data/ologit.dta")
head(dat)
```

---
## Example: Some Descriptive Stats

```{r}
## one at a time, table apply, pared, and public
lapply(dat[, c("apply", "pared", "public")], table)
```

---
## Example: Estimation
```{r}
library(MASS)
m <- polr(apply ~ pared + public + gpa, data = dat, Hess=TRUE)
## view a summary of the model
summary(m)
```

---
## Predicting probabailities

Let's say we want to know the probability that someone who's parents are educated is likely very likely to go to graduate school.

We want to know $$Pr(y=2|pared=1, public=0)$$

---
## Calculating probabilities
```{r}
## First let's find the cumulative probability of being less than 2.
gpa<-as.vector(seq(0,4,.1))
pared.f<-rep(1,length(gpa))
public.f<-rep(0,length(gpa))
fakedata=as.matrix(rbind(pared.f,public.f,gpa))
p2=exp(m$zeta[2]-t(fakedata)%*%m$coefficients)/(1+exp(m$zeta[2]
                +t(fakedata)%*%m$coefficients))
## but we want to know Pr(y=2)=1-Pr(y<2)
p2=1-p2
```

---
## Calculating probabilities
```{r echo=FALSE, message=FALSE, warning=FALSE, fig.align='center'}
## First let's find the cumulative probability of being less than 2.
library(tidyverse)
pared.f<-rep(0,length(gpa))
fakedata=as.matrix(rbind(pared.f,public.f,gpa))
p22=exp(m$zeta[2]-t(fakedata)%*%m$coefficients)/(1+exp(m$zeta[2]+t(fakedata)%*%m$coefficients))
## but we want to know Pr(y=2)=1-Pr(y<2)
p22=1-p22
g_data<-data.frame(p2, p22, gpa)
ggplot()+geom_line(data=g_data, aes(y=p22, x=gpa,color="pared=0"))+geom_line(data=g_data, aes( y=p2, x=gpa,color="pared=1"))+
xlab("Grade Point Average")+ylab("Prob. of Likely going to grad school")+labs(colour="Parents' Education")+theme_light()
```

---
class: inverse, center, middle
# Ordered Probit and Logit in R
---
## Marginal Effects

Re-run the example from the notes, but this time using the `oglmx` package

```{r message=FALSE, warning=FALSE}
library(erer)
m_probit <- polr(apply ~ pared + public + gpa, data = dat, Hess=TRUE, method = "probit")
m_logit <- polr(apply ~ pared + public + gpa, data = dat, Hess=TRUE, method = "logistic")
m1_probit<-ocME(m_probit,rev.dum = TRUE)
m1_logit<-ocME(m_logit,rev.dum = TRUE)
```

---
## Marginal Effects
```{r eval=TRUE, echo=FALSE}
m1_probit
m1_logit
```

---
class: inverse, center, middle
# Poisson and Negative Binomial Model
---
## Count Models

Count data are integer counts of the outcome variable that can include the number zero.

These data present problems for typical linear regression methods because the outcome variable is non-negative, but the regression error is bounded by $-\infty$ and $\infty$.

We cannot simply take a the natural log of the outcome variable because zero's are present.

Also, the integer values are more discrete in nature.

---
## OLS vs Count Models

These models have a number of advantages over an ordinary linear regression model, including a skew, discrete distribution, and the restriction of predicted values to non-negative numbers. A Poisson model is similar to an ordinary linear regression, with two exceptions. 

1. it assumes that the errors follow a Poisson, not a normal, distribution. 
2. rather than modeling Y as a linear function of the regression coefficients, it models the natural log of the response variable, ln(Y), as a linear function of the coefficients.
---
## Examples of count data

Count data capture the number of occurrences over a fixed time intervals such as:

- Number of Suicides in a state over a year
- Number of cars served through a fast food drive thru in a day
- Number of children in a family
- Number of doctors visits
- Number of website visitors

---
## Poisson Distribution

The Poisson density function is $$Pr(Y=y)=\frac{e^{-\lambda}\lambda^y}{y!}$$

where the $E[y|X]=Var[y|X]=\lambda$

For our purposes, we assume $\lambda=exp(X\beta)$

Interpretation of $\beta$ is the percentage change in $\lambda$ for a one unit increase in $X$.

[If you are completely unfamiliar with the Poisson distribution, then have a look at this video.](https://youtube.com/shorts/gSUv-l0HhD8?si=meKq-K8NlCQy-q_Z)

---
## Properties of the Poisson Distribution

The _equidispersion property_ of the Poisson distribution: Implies $$E[y|X]=Var[y|X]=\lambda$$, which means "the average number of occurrences is equal to the variance of occurrences". This assumption is very often violated.

When this property fails to hold, we state that there is _overdispersion_ in the data. In this case, the negative binomial can be used.

_Overdispersion_ causes the standard errors estimated using the Poisson distribution to be too small. An alternative approach is to estimate heterosckedastic (robust) standard errors. 

Woolridge (1999) had demonstrated that using robust standard errors compensates for the overdispersion problem and often performs better than the negative binomial distribution.

---
## Negative Binomial Distribution

The Negative Binomail distribution has a less restrictive relationship between the mean and variance of y. $$Var(y)=\lambda+\alpha \lambda^2$$

We can think of $\alpha$ as the _overdispersion parameter_.

---
## Testing for Overdispersion

Step 1. Estimate the count model using the negative binomial distribution.

Step 2. Test $H_o:\alpha=0$ vs $H_a:\alpha\neq0$

Step 3. Determine which case you are in

  * If $\alpha=0$, then the Poisson model holds
  * If $\alpha>0$, then we have _overdispersion_ (very common)
  * If $\alpha<0$, then we have _underdispersion_ (rare)

---
## Poisson and Negative Binomial in R
```{r fig.align='center', message=FALSE, warning=FALSE}
library(MASS); library(AER) #MASS contains the negative binomial model
data("NMES1988") 
plot(table(NMES1988$visits),xlab="Physician Office Visits", ylab="Counts")
```

---
## Issues with the data

First, notice the shape of Physician office visits. Does this look normally distributed to you?

Should we be concerned with the mass of zero's?

How about those outliers?

__89 visits is equivalent to a doctor's visit every 4 days!__

---
## Standard Poisson and Negative Binomial Model
```{r eval=FALSE}
pformu<-visits~hospital+chronic+insurance+gender+school
fm_pois <- glm(pformu, data = NMES1988, family = poisson)
cov.m1 <- vcovHC(fm_pois, type="HC0")
std.err <- sqrt(diag(cov.m1))
fm_qpois <- glm(pformu, data = NMES1988, family = quasipoisson) 
fm_nbin <- glm.nb(pformu, data = NMES1988)  
texreg::htmlreg(list(fm_pois,fm_pois,fm_qpois,fm_nbin),custom.model.names = c("Poisson","Poisson Robust","Quasi Poisson","Neg. Bin"), override.se = list(NULL,std.err,NULL,NULL), single.row = TRUE, custom.coef.names = c("Intercept","Hospital","Chronic","Insurance Eyes","School","Male","Theta"))
```

---
### Standard Poisson and Negative Binomial Model 

```{r echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
pformu<-visits~hospital+chronic+insurance+gender+school
fm_pois <- glm(pformu, data = NMES1988, family = poisson)
cov.m1 <- vcovHC(fm_pois, type="HC0")
std.err <- sqrt(diag(cov.m1))
fm_qpois <- glm(pformu, data = NMES1988, family = quasipoisson) 
fm_nbin <- glm.nb(pformu, data = NMES1988)  
texreg::htmlreg(list(fm_pois,fm_pois,fm_qpois,fm_nbin),custom.model.names = c("Poisson","Poisson Robust","Quasi Poisson","Neg. Bin"), override.se = list(NULL,std.err,NULL,NULL), single.row = TRUE, custom.coef.names = c("Intercept","Hospital","Chronic","Insurance Eyes","School","Male"))
```

---
### Fixed Effects Poisson

```{r eval=FALSE}
library(pglm)
library(readstata13)
library(lmtest)
library(MASS)
ships<-readstata13::read.dta13("http://www.stata-press.com/data/r13/ships.dta")
ships$lnservice=log(ships$service)
res <- glm(accident ~ op_75_79+co_65_69+co_70_74+co_75_79+lnservice,family = poisson, data = ships)
res1 <- pglm(accident ~ op_75_79+co_65_69+co_70_74+co_75_79+lnservice,family = poisson, data = ships, effect = "individual", model="within", index = "ship")
```

---
### Fixed Effects Poisson

```{r echo=FALSE ,warning=FALSE,message=FALSE}
library(pglm)
library(readstata13)
library(lmtest)
library(MASS)
ships<-readstata13::read.dta13("http://www.stata-press.com/data/r13/ships.dta")
ships$lnservice=log(ships$service)
res <- glm(accident ~ op_75_79+co_65_69+co_70_74+co_75_79+lnservice,family = poisson, data = ships)
res1 <- pglm(accident ~ op_75_79+co_65_69+co_70_74+co_75_79+lnservice,family = poisson, data = ships, effect = "individual", model="within", index = "ship")
coeftest(res)
standard_se<-ginv(-res1$hessian)
coeftest(res1,standard_se)
```

---
## Other Issues with Count Models

- Excessive zeros
  - Zero-inflated Models: function in R `zeroinfl()`
  - Hurdle Models: function in R `hurdle()`
- While many software packages implement standard count models, such as the Poisson and negative binomial models, more elaborate models may require some programming by the researcher.
- A count data approach does not solve the fundamental evaluation problem: absent a
randomized controlled experiment, identifying policy effects from observational data can be marred by selection bias, requiring plausibly exogenous variation in the form of a quasi-natural experiment
