---
title: "Limited Dependent Variables"
subtitle: "Share Regression"
author: "Jose M. Fernandez"
institute: "University of Louisville"
date: "2020-7-4 (updated: `r Sys.Date()`)"
output:
  xaringan::moon_reader:
    css: [default, uol, uol-fonts, "extra.css"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

### Logit with aggregate data

Logit has a nice mathematical property that allows us to estimate the model using both individual level and aggregate data. Recall, the logit probability is represented by 

$$Pr(Y=1|X)=F(X\beta)=\frac{exp(X\beta)}{1+exp(X\beta)}$$

Suppose you do not observe Y, but you do observed the fraction of people for where Y=1. Let's call this fraction, s.

Second, let's look at the formula for the log odds.

Odds is just the ratio of the two probabilities $\frac{Pr(Y=1)}{Pr(Y=0)}$.

log odds is the natural log of this fraction. $$log \space odds: ln \frac{Pr(Y=1)}{Pr(Y=0)}=ln \frac{s_1}{s_0}= ln(s_1)-ln(s_0)=X\beta$$.

---
### Natural log rules

In case it has been awhile since you have seen the natural log, here are some rules

- $ln x*y = ln x + ln y$
- $ln x/y = ln x - ln y$
- $ln x^a = a*ln x$
- $ln e = 1$
- $ln 1 = 0$



---
### Logit with aggregate data
The log odds gives us a linear function that we can estimate using OLS.

$$\begin{align} ln Pr(Y=1) &= ln F(X\beta)= ln \frac{exp(X \beta)}{1+exp(X \beta)} \\  &= ln\space exp(X\beta) - ln [1+exp(X \beta)] \\ &= X \beta - ln[1+exp(X \beta)] \end{align}$$
$$\begin{align} ln Pr(Y=0) &= ln [1-F(X\beta)]= ln \frac{1}{1+exp(X \beta)} \\  &= ln\space 1 - ln [1+exp(X \beta)] \\ &=  - ln[1+exp(X \beta)] \end{align}$$
Using these two equations we find $$ln Pr(Y=1)-ln Pr(Y=0)=X\beta \\ ln(s_1)-ln(s_0)=X\beta$$
---
### An Example

Let me give an example to better understand this advantage.

Suppose you have market shares for different products in the same class. Let $s_i$ be the market share of product $i$ and let $s_0$ be the percentage of people who do not buy a product in this class.

$$log(s_i)-log(s_0) = \beta_0 + \beta_1*price_i+\gamma*X_i+\epsilon_{ij}$$

We can estimate a demand function using market level data instead of individual level data. 

We would still need to find an instrumental variable for price.
---
### Let's Try it
```{r}
# Let's create individual level data for a logit
X <- rnorm(100000)*5+10
e1 <- rnorm(100000)*10
x1<-1:1000
w<-c(rep(1,100))
x2<-kronecker(x1,w)

B <- 0.7
y_star <- X*B + e1
y <- ifelse(y_star>0,1,0)
mean(y)
mydata <- data.frame(y,X,x2)
mylogit1 <- glm(y ~ X, family = "binomial", data = mydata)
mydata_group<-aggregate(mydata[, 1:2],list(mydata$x2), mean)
# now y is represented as a share
# Let's transform the shares using logs
mydata_group$y2 <- log(mydata_group$y)-log(1-mydata_group$y)
mylogit3 <- lm(y2 ~ X, mydata_group)
```

---
#### Results
<font size=2>
```{r echo=FALSE,message=FALSE,warning=FALSE, results='asis'}
stargazer::stargazer(mylogit1,mylogit3,type="html")
```
</font>